### 2018-03-11 paper/2-background.tex
Tweaks to background

Build systems automate the execution of simple repeatable tasks for individual
users, as well as for large organisations. There are software build systems,
such as \Make~\cite{feldman1979make}, \Shake~\cite{mitchell2012shake} and
\Bazel~\cite{bazel}, as well as various incremental calculation engines, such
as \Excel~\cite{advanced_excel}. In this section we use these four examples to
introduce main domain-specific notions and requirements. Other notable examples
of build systems and their relation to these four will be discussed
in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.

@@

\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we therefore use a simple and sensible
approximation to a real \Make that you might find on your machine.} was developed
more than 40 years ago to automatically build software libraries and executable
programs from source code. It uses \emph{makefiles} to describe tasks (often
referred to as \emph{build rules}) and their dependencies in a simple text form.
For example:

@@

\noindent
The above makefile lists three tasks: (i) compile a utility library comprising
files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
executing\footnote{In this example we treat \cmd{gcc} as a pure function for the
sake of simplicity. In reality, there are multiple versions of \cmd{gcc} and the
actual binary that is used to compile and link files is often also listed as a task
dependency.} the command \cmd{gcc -c util.c}, (ii) compile the main source file
\cmd{main.c} into \cmd{main.o}, and (iii) link object files \cmd{util.o} and
\cmd{main.o} into the executable \cmd{main.exe}. The makefile contains the
complete information about the \emph{task dependency graph}, which is shown in
Fig.~\ref{fig-make}(a).

@@

If the user modifies the sources of the utility library and runs \Make, it will
perform a \emph{full rebuild}, because all three tasks transitively depend on
the library, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if the
user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient: indeed,
the file \cmd{util.o} does not need to be rebuilt, since its inputs have not
changed, see Fig.~\ref{fig-make}(c). Note that if the dependency graph is
\emph{acyclic} then each task needs to be executed at most once. Cyclic task
dependencies are typically not allowed in build systems although there are rare
exceptions, see~\S\ref{sec-iterative-compute}.

The following property is essential for build systems, it is their raison d'\^etre.

\definition[Minimality]{A build system is \emph{minimal} if it executes tasks at
most once per build and only if they transitively depend on files that changed
since the previous build.}\label{def-minimal}
\vspace{2mm}

To achieve minimality \Make relies on two main ideas: (i) it uses \emph{file
modification time} to detect the files that changed\footnote{Technically, you
can fool \Make by altering the modification time of a file without changing its
content, e.g. by using the command \cmd{touch}. \Make is therefore minimal only
under the assumption that you do not do that.}, and (ii) it constructs a
complete task dependency graph from the information contained in the makefile
and executes tasks in a \emph{topological order}. See further details
in~\S\ref{sec-implementation-make}.

@@

\noindent
Here the cell \cmd{C1} controls which branch of the \cmd{IF} function is used
to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and this is
not known statically. One might say that the value of \cmd{C1} is statically
known in this example, but imagine that it is a result of a long computation
chain -- its value will then only become available during the build.

>>>>

Build systems automate the execution of simple repeatable tasks for individual
users, as well as for large organisations. There are software build systems,
such as \Make~\cite{feldman1979make}, \Shake~\cite{mitchell2012shake} and
\Bazel~\cite{bazel}, as well as various incremental calculation engines, such
as \Excel~\cite{advanced_excel}. In this section we use these four examples to
introduce main domain-specific notions and requirements. Other notable examples
of build systems and their relation to these four are discussed
in~\S\ref{sec-related} and~\S\ref{sec-conclusions}.

We start by giving an intuitive definition of a build system. After providing
more context and examples in the rest of this section, we
develop a formal definition in~\S\ref{sec-general-build}.

\definition[Build system]{Given a description of tasks, a \emph{build system}
executes them in an order that respects their dependencies to produce a
specified output from supplied inputs.}
\label{def-build}

@@

\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we therefore use a simple and sensible
approximation to a real \Make that you might find on your machine.} was developed
more than 40 years ago to automatically build software libraries and executable
programs from source code. It uses \emph{makefiles} to describe tasks (often
referred to as \emph{build rules}) and their dependencies in a simple textual form.
For example:

@@

\noindent
The above makefile lists three tasks: (i) compile a utility library comprising
files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
executing\footnote{In this example we pretend \cmd{gcc} is a pure function for the
sake of simplicity. In reality, there are multiple versions of \cmd{gcc} and the
actual binary that is used to compile and link files is often also listed as a task
dependency.} the command \cmd{gcc -c util.c}, (ii) compile the main source file
\cmd{main.c} into \cmd{main.o}, and (iii) link object files \cmd{util.o} and
\cmd{main.o} into the executable \cmd{main.exe}. The makefile contains the
complete information about the \emph{task dependency graph}, which is shown in
Fig.~\ref{fig-make}(a).

@@

If the user runs \Make specifying \cmd{main.exe} as the desired output, \Make
will first build \cmd{util.o} and \cmd{main.o}, in any order since the
corresponding tasks are independent, and then build \cmd{main.exe}. If the
user modifies the sources of \cmd{util.h} and runs \Make again, it will
perform a \emph{full rebuild}, because all three tasks transitively depend on
the \cmd{util.h}, as illustrated in Fig.~\ref{fig-make}(b). On the other hand, if the
user modifies \cmd{main.c} then a \emph{partial rebuild} is sufficient:
the file \cmd{util.o} does not need to be rebuilt, since its inputs have not
changed, see Fig.~\ref{fig-make}(c). Note that if the dependency graph is
\emph{acyclic} then each task needs to be executed at most once. Cyclic task
dependencies are typically not allowed in build systems although there are rare
exceptions, see~\S\ref{sec-iterative-compute}.

The following property is essential for build systems, it is their raison d'\^etre:

\definition[Minimality]{A build system is \emph{minimal} if it executes tasks at
most once per build and only if they transitively depend on inputs that changed
since the previous build.}\label{def-minimal}
\vspace{2mm}

To achieve minimality \Make relies on two main ideas: (i) it uses \emph{file
modification time} to detect which files changed\footnote{Technically, you
can fool \Make by altering the modification time of a file without changing its
content, e.g. by using the command \cmd{touch}. \Make is therefore minimal only
under the assumption that you do not do that.}, and (ii) it constructs a
complete task dependency graph from the information contained in the makefile
and executes tasks in a \emph{topological order}. For a more concrete description
see~\S\ref{sec-implementation-make}.

@@

\noindent
Here the cell \cmd{C1} controls which branch of the \cmd{IF} function is used
to compute \cmd{B1}. When \cmd{C1=1}, the dependencies of \cmd{B1} are
$\{\cmd{C1},\cmd{A1}\}$, otherwise they are $\{\cmd{C1},\cmd{A2}\}$, and which is
not known statically. One might say that the value of \cmd{C1} is statically
known in this particular example, but imagine that it is the result of a long computation
chain -- its value will only become available during the build.
